Correlation among smart grid actors

### 2015-05-11 Article.tex
update

@@

This problem exhibits similarities with our objective of forming stable coalitions since our expected return can be formulated as the expected production and the risk / volatility as the variance of the production. However, one of the assumptions in the portfolio theory is that returns are normaly distributed random variables (or at least that the joint distribution is elliptic). This is a strong assumption which reveals unrealistic in most cases for our purposes when looking at real data.

Nevertheless, one of the key point in the Markowitz theory is to consider explicitely the correlation structure between the assets since these correlation relationships impact directly (as we will see in section X) the variances of the portfolios. Since the work of Mantanega, an interresting approach consists in computing a distance metric based on the correlation coefficients in order to organize the series in a correlation graph where the weight of an edge between two timeseries (nodes) is the metric value for these series.

@@

An essential component of the smart grid is the smart meter which makes the interface between the end user and the rest of the system. Smart meters coupled with sensors measure quantities of interest (like instantaneous consumption), receive informations from the grid (electricity price for instance), and take actions accordingly (demand side management program). Smart meters are currently and gradually deployed, and will probably provide interesting datasets to work on. Unfortunately, at the time this paper was written, production and consumption data for prosumers over a large region were not yet available to our knowledge. Some interesting experiments are notwithstanding being conducted and data are progressively made public (see for instance the ISSDA experiment in Ireland).

In the following, we denote by $ P_{i}(t) $ the instantaneous extra-production of agent i at time t. That is, $ P_{i}(t) = P_{i}^{P}(t) - P_{i}^{D}(t) $, where $ P_{i}^{P}(t) $ represents the total production of agent i at time t and $ P_{i}^{D}(t) $ its consumption at time t. In other words, $ P_{i}(t) $ represents the instantaneous surplus of power that agent i is willing to sell at time t. As explained above, since large datasets containing this quantity over time are not yet available, we simulated these traces by considering separately $ P_{i}^{P} $ and $ P_{i}^{D} $.

@@

Weather quantities like wind speed or solar irradiance appear thus as alternative data for generating the $ P_{i} $ series. Fortunately, these kind of data are easier to find, and since the developement of small personal weather stations, their geographical granularity keeps increasing.  A quantity like wind speed depends both on time and location. In the following, time is discretized into slots and space into zones. A zone is simply a portion of the considered space for which we have weather data. Therefore, if prosumers i and j are positioned on the same zone, they are exposed to the same weather. Adding some noise can easily be done though not considered in this paper. The process for generating the $ P_{i} $ series is pictured in the first block of the process diagram (see figure \ref{fig:process}).

Note that a prosumer i is defined by his zone $ Z_{i} $ as well as the sets $ DER_{i} $ and $ load_{i} $. That is, a prosumer can be configured to represent anything from a single windturbine for instance ($ DER_{i} = \{ WT_{0} \} $ and $ load_{i} = \emptyset $) to a pure load ($ DER_{i} = \emptyset $ and $ load_{i} = \{ L_{0} \} $) throug more complex combinations. In practice, we use random configurations for the agents.

@@

As explained in section \ref{sec:data}, we have a set $ \mathcal{A} = \{a_{1},a_{2},...,a_{N} \} $ of N prosumers configured randomly, and for each agent, we simulated his extra-production $ P_{i}(t),\ \forall i \in \mathcal{A} $ from 2006 to 2012. Based on these historical values, our objective is now to form groups of prosumers (the so-called coalitions) so that the global power production resulting from the superposition of individual's extra-productions be both sufficiently high and predictable. Let $ P_{S}(t) = \sum_{i \in S} P_{i}(t) $ be the extra-production of coalition S at time t.

@@

Let us illustrate the rest of the notations and concepts with a simple example. We consider only two agents i and j such that the distribution of their extra-production can be approximated by normal distributions : $ P_{i} \sim \mathcal{N}(\mu_{i}, \sigma_{i} ) $ and $ P_{j} \sim \mathcal{N}(\mu_{j}, \sigma_{j} ) $. This is only for explanation purposes as it is of course rather unrealistic in real situations where the distributions are skewed. Using simple statistics, we can write the distribution of the coalition $ \{i,j\} $ as $ P_{\{i,j\}} \sim \mathcal{N}(\mu_{ij}, \sigma_{ij}) $, where :

@@

$ \rho_{ij} $ being the Pearson's correlation coefficient between $ P_{i} $ and $ P_{j} $. If the coalition $ \{i,j\}$ proposes a contract value $ P_{S}^{CRCT} $, all instants where $ \{i,j\}$ will produce less than $ P_{S}^{CRCT} $ is critical. Indeed, in this situation, $ \{i,j\}$ will either have to discharge batteries to keep up with its contract, or pay penalities to the grid. The probability that $ \{i,j\}$ is underproducing compared to the contract : $ Pr[P_{i,j} \leq P^{CRCT}] $ is thus an important indicator of the coalition's quality. A well-known result for normal distributions is that the cumulative distribution function can be written as :
\begin{equation}
Pr[P_{ij} \leq P^{CRCT}] = \dfrac{1}{2} \left[ 1+ erf \left( \dfrac{P^{CRCT} - \mu_{ij}}{\sigma_{ij}\sqrt{2}} \right) \right] 
\end{equation}

@@

The amount of risks a given coalition is willing to take depends on a lot of things, among wich its capacity to compensate for under-producing (using batteries, backup generators...). Selecting the right contract value appears thus as an interesting problem on its own that we plan to investigate in future works. In order to keep the present paper in a reasonable length, we simplify the contract value selection problem a little bit by giving some responsabilities to a third party named the grid operator. The role of the grid operator is to constrain the market entry to coalitions able to propose both sufficiently high and sufficiently credible contract values. More formally, let $ \phi \in [0,1] $ be a threshold fixed by the grid operator as a maximum value for the probability of under-producing. The highest contract value that a coalition can propose is thus $ P_{S}^{CRCT \star} $ such that $ Pr[P_{ij} \leq P_{S}^{CRCT \star}] = \phi $. In the gaussian example, this translates by coalition $ \{i,j\}$ announcing :

\begin{equation}
\label{eq:P_star}
P_{S}^{CRCT \star} = \mu_{ij} + \sqrt{2} \sigma_{ij} erf^{-1}(2 \phi - 1 )
\end{equation}

@@

This is the best contract value that the coalition S can afford giving the stability policy $\phi$ of the grid operator. Figure \ref{fig:Gaussian} shows how $ P_{S}^{CRCT \star} $ evolves according to the reliability parameter $ \phi $. For illustration, the range of $ \phi $ values is shown from 0 to 1, but in practice, only small values of $ \phi $ really make sense : $ \phi = 1 $ for instance means that coalitions can announce absolutely anything since the probability of producing less than any contract value is necessarily less than one by trivial definition of a probability. As visible on figure \ref{fig:Gaussian}, coalitions with high expected production but presenting a high unpredictability are penalized and can only afford small contracts.

@@

Before running the algorithm, we need to calibrate the utility function by choosing the $ \alpha $ parameter. Recall that the purpose of this parameter is to take into account some constraints on the coalition's sizes if needed. In this paper, neither the communication network nor the electrical grid are explicitely considered. Thus, we do not have any technical constraints on coalition sizes even if we designed the utitility such that these could be taken into account. We select the desired size as being $ \lfloor N/N_{COAL} \rfloor $ (where $ \lfloor.\rfloor $ means floor). Figure \ref{fig:real_utility2} shows how the mean utility of a coalition evolves with its size when the optimum size is set to 40 agents. Using equation \ref{eq:alpha_star} to select $ \alpha $ based on the mean quantities seems to give acceptable results for the utility function.

@@

Figure \ref{fig:coalitions} shows the coalitions formed with the considered algorithms. A coalition is represented by a marker in the contract value / volatility space, and the color and shape of a marker indicates by which algorithm the coalition has been formed. Besides, the size of a marker is proportional to the coalition size in number of agents. The coalitions of correlated agents are clearly of poor quality according to our criteria since they can only afford small production's contracts, and with a very high volatility (compared to the other coalitions on the plot). On figure \ref{fig:coalitions}, the de-correlated coalitions and the randomized ones are closer to the bottom right corner indicating a much better quality. We can see that the de-correlated coalitions are globally more located toward the bottom right corner, meaning that there are less disparities between coalitions than with the random case. Besides, we can see that the coalitions seem quite balanced in term of the number of agents that they contain.

@@

>>>>

Because agents are susceptible to fail for diverse reasons, the impact on the system's stability is critical. The propensy of a system to undertake these failures is often refered to as its resilience. Once formed, we also investigate in this paper the resilience of the coalitions when prosumers fail. Despite the fact that loosing agents is always detrimental to the coalitions, we will see that, with our algorithm, coalitions tend to be less impacted by random failures.

@@

This problem exhibits similarities with our objective of forming stable coalitions since our expected return can be formulated as the expected production and the risk / volatility as the variance of the production. However, one of the assumptions in the portfolio theory is that returns are normaly distributed random variables (or at least that the joint distribution is elliptic). In this paper, we chose to use normal distributions only for explanation purposes, or for estimating some parameters. Therefore, all the results are computed based on observed distributions (sometines far from gaussians).

One of the key point in the Markowitz theory is to consider explicitely the correlation structure between the assets since these correlation relationships impact directly (as we will see in section X) the variances of the portfolios. Since the work of Mantanega, an interresting approach consists in computing a distance metric based on the correlation coefficients in order to organize the series in a correlation graph where the weight of an edge between two timeseries (nodes) is the metric value for these series.

@@

An essential component of the smart grid is the smart meter which makes the interface between the end user and the rest of the system. Smart meters coupled with sensors measure quantities of interest (like instantaneous consumption), receive informations from the grid (electricity price for instance), and take actions accordingly (demand side management program). Smart meters are currently and gradually deployed, and will probably provide interesting datasets to work on. Unfortunately, at the time this paper was written, production and consumption data for prosumers over a large region were not yet available to our knowledge. Some interesting experiments are notwithstanding being conducted and data are progressively made public [].

In the following, we denote by $ P_{i}(t) $ the instantaneous extra-production of agent i at time t :

\begin{equation}
P_{i}(t) = P_{i}^{P}(t) - P_{i}^{D}(t)
\end{equation}

Where $ P_{i}^{P}(t) $ represents the total production of agent i at time t and $ P_{i}^{D}(t) $ its consumption at time t. In other words, $ P_{i}(t) $ represents the instantaneous surplus of power that agent i is willing to sell at time t. As explained above, since large datasets containing this quantity over time are not yet available, we simulated these traces by considering separately $ P_{i}^{P} $ and $ P_{i}^{D} $.

@@

Weather quantities like wind speed or solar irradiance appear thus as alternative data for generating the $ P_{i} $ series. Fortunately, these kind of data are easier to find, and since the developement of small personal weather stations, their geographical granularity keeps increasing.  A quantity like wind speed depends both on time and location (we neglect height since this information is rarely present in the data). In the following, time is discretized into slots and space into zones. A zone is simply a portion of the considered space for which we have weather data. Therefore, if prosumers i and j are positioned on the same zone, they are exposed to the same weather. Adding some intra-zone noise can easily be done though not considered in this paper. The process for generating the $ P_{i} $ series is pictured in the first block of the process diagram (see figure \ref{fig:process}).

Note that a prosumer i is defined by his zone $ Z_{i} $ as well as the sets $ DER_{i} $ and $ load_{i} $. That is, a prosumer can be configured to represent anything from a single windturbine for instance ($ DER_{i} = \{ WT_{0} \} $ and $ load_{i} = \emptyset $) to a pure load ($ DER_{i} = \emptyset $ and $ load_{i} = \{ L_{0} \} $) through more complex combinations. In practice, we use random configurations for the agents.

@@

This section provides most of the notations and introduces important concepts for the rest of the paper. As explained in section \ref{sec:data}, we consider a set $ \mathcal{A} = \{a_{1},a_{2},...,a_{N} \} $ of N prosumers configured randomly, and for each agent, we simulate its extra-production $ P_{i}(t),\ \forall i \in \mathcal{A} $ from 2006 to 2012. Based on these historical values, our objective is now to form groups of prosumers (the so-called coalitions) so that the global power production resulting from the superposition of individual's extra-productions be both sufficiently high and predictable. Let $ P_{S}(t) = \sum_{i \in S} P_{i}(t) $ be the extra-production of coalition S at time t.

@@

Let us illustrate the rest of the notations and concepts with a simple example. We consider only two agents i and j such that the distribution of their extra-production can be approximated by normal distributions : $ P_{i} \sim \mathcal{N}(\mu_{i}, \sigma_{i} ) $ and $ P_{j} \sim \mathcal{N}(\mu_{j}, \sigma_{j} ) $. This is only for explanation purposes as it is of course rather unrealistic in real situations where the distributions are skewed. Using simple statistics, we can write the distribution of the coalition $ S = \{i,j\} $ as $ P_{\{i,j\}} \sim \mathcal{N}(\mu_{ij}, \sigma_{ij}) $, where :

@@

$ \rho_{ij} $ being the Pearson's correlation coefficient between $ P_{i} $ and $ P_{j} $. If the coalition $ \{i,j\}$ proposes a contract value $ P_{S}^{CRCT} $, all instants where $ \{i,j\}$ will produce less than $ P_{S}^{CRCT} $ is critical. Indeed, in this kind of situations, $ \{i,j\}$ will either have to discharge batteries to keep up with its contract, or pay penalities to the grid. The probability that $ \{i,j\}$ is underproducing compared to the contract : $ Pr[P_{i,j} \leq P^{CRCT}] $ is thus an important indicator of the coalition's quality. A well-known result for normal distributions is that the cumulative distribution function can be written as :
\begin{equation}
Pr[P_{ij} \leq P_{S}^{CRCT}] = \dfrac{1}{2} \left[ 1+ erf \left( \dfrac{P_{S}^{CRCT} - \mu_{ij}}{\sigma_{ij}\sqrt{2}} \right) \right] 
\end{equation}

@@

The amount of risks a given coalition is willing to take depends on a lot of things, among wich its capacity to compensate for under-producing (using batteries, backup generators...), or the risk aversion of the aggregator. Selecting the right contract value appears thus as an interesting problem on its own that we plan to investigate in future works. In order to keep the present paper in a reasonable length, we simplify the contract value selection problem a little bit by giving some responsabilities to a third party named the grid operator. The role of the grid operator is to constrain the market entry to coalitions able to propose both sufficiently high and sufficiently credible contract values. More formally, let $ \phi \in [0,1] $ be the reliability threshold fixed by the grid operator as a maximum value for the probability of under-producing. The highest contract value that a coalition can propose is thus $ P_{S}^{CRCT \star} $ such that $ Pr[P_{ij} \leq P_{S}^{CRCT \star}] = \phi $. In the gaussian example, this translates by coalition $ \{i,j\}$ announcing :

\begin{equation}
\label{eq:P_star}
P_{S}^{CRCT \star} = \mu_{ij} - \sqrt{2} \sigma_{ij} erf^{-1}( 1 - 2 \phi )
\end{equation}

@@

This is the best contract value that the coalition S can afford giving the stability policy $\phi$ of the grid operator. Figure \ref{fig:Gaussian} shows how $ P_{S}^{CRCT \star} $ evolves according to the reliability parameter $ \phi $. For illustration, the range of $ \phi $ values is shown from 0 to 1, but in practice, only small values of $ \phi $ really make sense : $ \phi = 1 $ for instance means that coalitions can announce absolutely anything since the probability of producing less than any contract value is necessarily less than one by trivial definition of a probability. As visible on figure \ref{fig:Gaussian}, coalitions with high expected productions but presenting a high unpredictability are penalized and can only afford small contracts.

@@

Before running the algorithms, we need to calibrate the utility function by choosing the $ \alpha $ parameter. Recall that the purpose of this parameter is to take into account some constraints on the coalition's sizes if needed. In this paper, neither the communication network nor the electrical grid are explicitely considered. Thus, we do not have any technical constraints on coalition sizes even if we designed the utitility such that these could be taken into account. We select the desired size as being $ \lfloor N/N_{COAL} \rfloor $ (where $ \lfloor.\rfloor $ means floor). Figure \ref{fig:real_utility2} shows how the mean utility of a coalition evolves with its size when the optimum size is set to 40 agents. Using equation \ref{eq:alpha_star} to select $ \alpha $ based on the mean quantities seems to give acceptable results for the utility function.

@@

Figure \ref{fig:coalitions} shows the coalitions formed with the considered algorithms. A coalition is represented by a marker in the contract value / volatility space, and the color and shape of a marker indicates by which algorithm the coalition has been formed. Besides, as the size of a marker is proportional to the number of agents in the coalition, we can see that the utility function results in approximatively balanced coalitions. The coalitions of correlated agents (green squares) are clearly of poor quality according to our criteria since they can only afford small production contracts, and with a very high volatility (compared to the other coalitions on the plot).

On figure \ref{fig:coalitions}, the de-correlated coalitions (blue dots) and the random ones (red triangles) are closer to the bottom right corner indicating a much better quality. We can see that, despite its simplicity, the random algorithm is able to achieve correct results thanks to the diversification inherent to such algorithm. Nevertheless, based on the utility, coalitions resulting from this random search are less good than the de-correlated ones. This can be understood on figure \ref{fig:coalitions} as the de-correlated coalitions are globally more located toward the bottom right corner, meaning that there are less disparities between coalitions than with the random case.

@@

A key point for the coalitions, besides stability and productivity, is their resilience. The resilience of a system can be roughly described as its ability to perform its tasks when subject to failures of its components. Therefore, the notion of resilience we will use in the following can be seen as the ability of the coalition structures to inject stable power in the grid when node failures occur. According to our model, the grid operator specified two thresholds ($P^{MIN}$ and $ \phi $) such that the power injected by every coalition is constrained : $ P_{S}^{CRCT} \in [P^{MIN}, P_{S}^{CRCT \star}] $. As long as a coalition can propose a contract value higher than $ P^{MIN} $, it is valid and allowed to enter the energy market. We define the resilience of a coalition S as the probability that S produces more than the $ P^{MIN} $ threshold :

\begin{equation}
\mathcal{R}_{S} = 1 - Pr[P_{S} < P^{MIN}]
\end{equation}

And we extend this measure to the coalition stuctures :

\begin{equation}
\mathcal{R}_{CS} = \prod_{S \in CS} \left( 1 - Pr[ P_{S} < P^{MIN} ] \right)
\label{eq:resilience}
\end{equation}

We consider that prosumers fail randomly, and we denote by $ \psi \in [0,1] $ the fraction of agents that failed. Figure \ref{fig:resilience} exhibits how the resilience of the coalition structures evolve according to $ \psi $. On the top subplot, $ P^{MIN} $ was volontarily selected relatively low such that the resiliences of the three structures fit on the same figure. When the $ P^{MIN} $ requirement increases, the differences between the algorithms also increase as visible on the bottom subplot of figure \ref{fig:resilience}. The de-correlated coalitions seem to achieve a more resilient production on the market in the sense that they are able to sustain a higher fraction of node failures.

\begin{figure}
\includegraphics[scale=.6]{./figures/resilience_both.pdf}
\caption{Resilience of the coalition structures when nodes fail randomly (see equation \ref{eq:resilience}) for $ P^{MIN} = 10MW $ (top subplot) and $ P^{MIN} = 80MW $ (bottom subplot) }
\label{fig:resilience}
\end{figure}
